---
title: 'CAGE: Check AEGIS1 analysis'
author: "John Thompson"
date: "2023-03-29"
output: 
    prettydoc::html_pretty:
      toc: true
      theme: tactile
---

```{r setup, include=FALSE}
library(tidyverse)
library(glue)
library(fs)
library(ggthemes)

cache <- "C:/Projects/RCourse/Masterclass/CAGE/data/cache"

# --- source display functions
code <- "C:/Projects/RCourse/Masterclass/CAGE/code"
source(path(code, "display_functions.R"))

# --- read data from cache
cache <- "C:/Projects/RCourse/Masterclass/CAGE/data/cache"
readRDS(path(cache, "patients.rds")) %>%
  mutate(class = ifelse(diagnosis == "Cancer", 1, 0)) %>% 
  select(id, diagnosis, class) -> classDF
  

readRDS(path(cache, "aegis1_loss.rds")) -> loss


knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center",
                      message=FALSE)

theme_set(theme_stata())
```

## Background

Results for feature selection from the full set of probes taken from aegis1.R

## Warning Messages

```{r}
readLines(path(cache, "aegis1_log.txt"))
```

## Unscaled PCA

```{r}
vpca <- readRDS(path(cache, "aegis1_vpca.rds"))

# ---------------------------------------------
# Principal Components Stdev (root eigenvalues)
#
plot_eigenvalues(vpca$sdev) +
  ggtitle("Standard deviatons of the Principal Components")
```

## Unscaled PCA scores training data

```{r}
readRDS(path(cache, "aegis1_vpca_scores.rds")) %>%
  print() -> vScoreTrainDF
# ---------------------------------------------
# Illustrative plot of first 5 PCs by diagnosis
# training data
#
glue("PC{1:5}") %>%
  boxplot_features(vScoreTrainDF %>%
                      left_join(classDF, by = "id")) +
  labs(title = "Training data: PCA Scores by diagnosis",
       y     = "Score",
       x     = "Principal Component")
```

## Unscaled PCA scores validation Data

```{r}
readRDS(path(cache, "aegis2_vpca_scores.rds")) %>%
  print() -> vScoreValidDF

# ---------------------------------------------
# Illustrative plot of first 5 PCs by diagnosis
# validation data
#
glue("PC{1:5}") %>%
  boxplot_features(vScoreValidDF %>%
                      left_join(classDF, by = "id")) +
  labs(title = "Validation data: PCA Scores by diagnosis",
       y     = "Score",
       x     = "Principal Component")
```

## Scaled PCA

```{r}
cpca <- readRDS(path(cache, "aegis1_cpca.rds"))

# ---------------------------------------------
# Principal Components Stdev (root eigenvalues)
#
plot_eigenvalues(cpca$sdev) +
  ggtitle("Standard deviatons of the Principal Components")
```

## Scaled PCA scores training data

```{r}
readRDS(path(cache, "aegis1_cpca_scores.rds")) %>%
  print() -> cScoreTrainDF
# ---------------------------------------------
# Illustrative plot of first 5 PCs by diagnosis
# training data
#
glue("PC{1:5}") %>%
  boxplot_features(cScoreTrainDF %>%
                      left_join(classDF, by = "id")) +
  labs(title = "Training data: PCA Scores by diagnosis",
       y     = "Score",
       x     = "Principal Component")
```

## Scaled PCA scores validation Data

```{r}
readRDS(path(cache, "aegis2_cpca_scores.rds")) %>%
  print() -> cScoreValidDF

# ---------------------------------------------
# Illustrative plot of first 5 PCs by diagnosis
# validation data
#
glue("PC{1:5}") %>%
  boxplot_features(cScoreValidDF %>%
                      left_join(classDF, by = "id")) +
  labs(title = "Validation data: PCA Scores by diagnosis",
       y     = "Score",
       x     = "Principal Component")
```

## Best Probes

```{r}
loss$probeUniDF %>%
  print() %>%
  arrange(loss) %>%
  print()
```

```{r}
print(loss$probeSelectedDF)
```


## Unscaled PCA: Most Predictive PCs

```{r}
loss$vpcaUniDF %>%
  print() %>%
  arrange(loss) %>%
  print()

loss$vpcaSelectedDF %>%
  print()
```

```{r}
loss$vpcaOrderedDF %>%
  print()
```

## Unscaled PCA: First PCs

```{r}
loss$vpcaOrderedDF %>%
  print()
```

## Scaled PCA: Most Predictive PCs

```{r}
loss$cpcaUniDF %>%
  print() %>%
  arrange(loss) %>%
  print()

loss$cpcaSelectedDF %>%
  print()
```

## Scaled PCA: First PCs

```{r}
loss$cpcaOrderedDF %>%
  print()
```
